---
title: 'refitME: Tutorial for fitting MCEM models when covariates are subject to measurement error/error-in-variables'
author: "Jakub Stoklosa & David Warton"
date: "18th Feb 2019"
header-includes: # Allows you to add in your own Latex packages.
- \usepackage{float} # Use the 'float' package.
- \floatplacement{figure}{H} # Make every figure with caption = h
- \usepackage{ragged2e}
- \usepackage{geometry}
- \geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
output:
  html_document: default
  fig_caption: yes
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

This tutorial documents fitting an MCEM algorithm via the `refitME` R-package. For more specific details see:

Stoklosa, J. and Warton, D.I. (2019). A general algorithm for error-in-variables using Monte Carlo expectation maximization.

Also, see ```?refitME``` for futher details on the fitting function, input arguments and output.

## Example 1: A simple GLM example taken from Carroll _et al._ (2006). 

The Framingham heart study data set. Here, we also fit SIMEX models and compare them with MCEM. Computationanl times for both models are also reported.

### Load data and R-packages.

```{r, echo=TRUE}
suppressMessages(library(refitME));
suppressMessages(library(simex));

epsilon<-0.00001;  # A set convergence threshold.
B<-100;  # The number of Monte Carlo replication values/SIMEX simulations.

family<-"binomial";

data.Fram<-as.matrix(read.table(file="Framinghamdata.txt"));
```

### Setup all variables (the construction below follows the Carroll et al. (2006) monograph).

```{r, echo=TRUE}
Y<-data.Fram[,10]; # Binary variable.

n<-length(Y);

z1<-(data.Fram[,9]); # Cholesterol.
z2<-(data.Fram[,2]); # Age.
z3<-data.Fram[,7];   # Smoke.
w1<-(log((data.Fram[,3]+data.Fram[,4]+data.Fram[,5]+data.Fram[,6])/4-50)); # Mean exam 2 and 3.
dat<-data.frame(cbind(Y,z1,z2,z3,w1));

sigma.sq.u<-0.01259/2 # ME variance, obtained from Carroll et al. (2006) monograph.
```

### Fit the naive model.

The first stored variable `w1` is the error contaminated variable in the analysis.

```{r, echo=TRUE}
mod_naiv1<-glm(Y~w1+z1+z2+z3,x=TRUE,family=binomial,data=dat); 
```

### Fit the SIMEX model.

```{r, echo=TRUE}
start<-Sys.time();
mod_simex1<-simex(mod_naiv1,SIMEXvariable=c("w1"),
                  measurement.error=cbind(sqrt(sigma.sq.u)),B=B); # SIMEX.
end<-Sys.time();
t1<-difftime(end,start,units="secs");
comp.time<-c(t1);
```

### Fit the MCEM model.

```{r, echo=TRUE}
start<-Sys.time();
est<-refitME(mod_naiv1,sigma.sq.u,B);
end<-Sys.time();
t2<-difftime(end,start,units="secs");
comp.time<-c(comp.time,t2);
```

### Report and compare times and model estimates.

```{r, echo=TRUE}
est.beta<-rbind(coef(mod_naiv1),coef(mod_simex1),est$beta);
est.beta.se<-rbind(sqrt(diag(vcov(mod_naiv1))),
                   sqrt(diag(mod_simex1$variance.jackknife)),est$beta.se2);
round(est.beta,digits=3);
round(est.beta.se,digits=3);  # Standard error estimates.
comp.time;  # SIMEX and MCEM.
```

\clearpage

## Example 2: A GAM example taken from Ganguli _et al._ (2005).

The Milan mortality air pollution data set. Here, we fit GAM models via the `mgcv` package where one covariate is error-contaminated.

### Load data and R-packages.

```{r, echo=TRUE}
rm(list=ls());

suppressMessages(library(refitME));
suppressMessages(library(SemiPar));

epsilon<-0.00001;  # A set convergence threshold.
B<-2;  # The number of Monte Carlo replication values.

family<-"poisson";

data(milan.mort);

dat.air<-milan.mort;
```

### Setup all variables.

```{r, echo=TRUE}
Y<-dat.air[,6];  # Mortality counts.

n<-length(Y);

z1<-(dat.air[,1]);
z2<-(dat.air[,4]);
z3<-(dat.air[,5]);
w1<-log(dat.air[,9]);
w1<-scale(w1);
colnames(w1)<-"w1";
dat<-data.frame(cbind(Y,z1,z2,z3,w1));

## Reliability ratio.

sigma.sq.u<-0.1; # Rel. ratio of 0.9.
#sigma.sq.u<-0.2; # Rel. ratio of 0.8.
#sigma.sq.u<-0.3; # Rel. ratio of 0.7.

rel.rat<-(1-sigma.sq.u/var(dat$w1))*100;
```

### Fit the naive model.

```{r, echo=TRUE}

mod_naiv1<-gam(Y~s(w1,k=5)+s(z1,bs='cc',k=25)+s(z2,k=5)+s(z3,k=5),family="poisson",data=dat);
```

### Fit the MCEM model.

```{r, echo=TRUE}
est<-refitME(mod_naiv1,sigma.sq.u,B);
```

```{r fig1, fig.height=7, fig.pos='H', echo=FALSE, fig.cap="\\label{fig:fig1} Plots of smooths against covariate. TSP (top left is the error contaminated variable)."}
xlab.names<-c("log(TSP)","Day","Temp","Humidity");

op<-par(mfrow=c(2,2),las=1);

for(i in 1:4)
  {
  if(i==1)
    {
    plot(est$mod,select=i,ylim=c(-0.4,0.2),rug=FALSE,col=3,all.terms=TRUE,
         xlab=xlab.names[i],ylab="s(Mortaity counts)",lwd=3,cex.lab=1,cex.axis=1,
         cex.main=2,font.lab=1,cex=1);
    title(main=bquote("Reliability ratio for log(TSP) = "~.(rel.rat)~"%"),
          outer=F,line=1,cex=1.4);
    for(j in 1:2)
      {
      axis(j,labels=FALSE);
      }
    }
  if(i==2)
    {
    plot(est$mod,select=i,ylim=c(-0.25,0.3),rug=FALSE,col=3,all.terms=TRUE,
         xlab=xlab.names[i],ylab="s(Mortaity counts)",lwd=3,cex.lab=1,cex.axis=1,
         cex.main=2,font.lab=1,cex=1);
    for(j in 1:2)
      {
      axis(j,labels=FALSE);
      }
    }
  if(i==3)
    {
    plot(est$mod,select=i,ylim=c(-0.2,0.4),rug=FALSE,col=3,all.terms=TRUE,
         xlab=xlab.names[i],ylab="s(Mortaity counts)",lwd=3,cex.lab=1,cex.axis=1,
         cex.main=2,font.lab=1,cex=1);
    for(j in 1:2)
      {
      axis(j,labels=FALSE);
      }
    }
  if(i==4)
    {
    plot(est$mod,select=i,ylim=c(-0.3,0.08),rug=FALSE,col=3,all.terms=TRUE,
         xlab=xlab.names[i],ylab="s(Mortaity counts)",lwd=3,cex.lab=1,cex.axis=1,
         cex.main=2,font.lab=1,cex=1);
    for(j in 1:2)
      {
      axis(j,labels=FALSE);
      }
    }
  }

title(main="MCEM (Poisson GAM) fitted to the air pollution data.",outer=T,line=-2);
par(op);
```
